Effectful Semantics for Substructural SSA

### 2023-11-16 pldi24/paper.tex
Tap

\TODO{\textit{what} is the semantics of a term}

We can now give a semantics for terms in Figure \ref{fig:ssa-term-semantics}.
Our treatment of \rle{var} is quite standard: the expression \(x\) types in
\(\Gamma\) if all other variables in \(\Gamma\) and can be discarded, in which
case the semantics is simply doing so, i.e.
\(\dnt{\cwk{\Gamma}{\thyp{x}{A}{q}}}\), lifting to \(\mc{C}_0\) as necessary.
Similarly, to type a function application \rle{var}, we first take
\(\dnt{\hasty{\Gamma}{1}{a}{A}{q}}\), which is guarnateed to be central, lift it
to the appropriate category \(\mc{C}_p\), and then simply postcompose it with
the denotation of \(f\). \rle{pair} is handled by composing the context split
\(\csplits{\Gamma}{\Delta}{\Xi}\) taking the tensor product of the components
\(\dnt{\hasty{\Delta}{1}{a}{A}{q}}\), \(\dnt{\hasty{\Xi}{1}{b}{B}{q}}\); this is
well-defined since both components are central and \(\mc{C}_1\) is monoidal. We
then lift the result to the appropriate \(\mc{C}_p\). The semantics of
\rle{unit} is simply dropping the entire context; \(\dnt{\cwk{\Gamma}{\cdot}}\),
while \rle{true}, and \rle{false} are each handled by first dropping the context
and then postcomposing with the appropriate constant; in all cases, this is
followed by lifting to \(\mc{C}_p\). Finally, \rle{let} and \rle{let2} are
handled by first splitting the context, then passing the right hand component to
the semantics of the bound expression \(\dnt{\hasty{\Xi}{1}{a}{A}{q}}\) or
\(\dnt{\hasty{\Xi}{1}{a}{A \otimes B}{q}}\), and then, lifting to \(\mc{C}_p\)
as necessary, after reassociating with \(\alpha\), passing everything to the
inner expression \(\dnt{\hasty{\Delta, \thyp{x}{A}{q}}{p}{e}{B}{q}}\) or
\(\dnt{\hasty{\Delta, \thyp{x}{A}{q}, \thyp{y}{B}{q}}{p}{e}{C}{q}}\). Note that
this inner expression is not necessarily central, unlike the bound expression!

>>>>

We can now interpret a term \(\hasty{\Gamma}{p}{a}{A}{q}\) as a
\textit{morphism} taking \(\dnt{\Gamma}\) into \(\dnt{A}\) with
\textit{centrality} \(p\) and \textit{quantity} \(q\) (i.e., lying in
\(\mc{C}_p^q\)). This means that, as expected, a central term becomes a central
morphism in \(\mc{C}_1\), an affine term becomes an affine morphism, and a
relevant term becomes a relevant morphism, and so, in particular, a pure term
becomes a pure morphsim. Our rules for doing so are given in Figure
\ref{fig:ssa-term-semantics}. Our treatment of \rle{var} is quite standard: the
expression \(x\) types in \(\Gamma\) if all other variables in \(\Gamma\) and
can be discarded, in which case the semantics is simply doing so, i.e.
\(\dnt{\cwk{\Gamma}{\thyp{x}{A}{q}}}\), lifting to \(\mc{C}_0\) as necessary.
Similarly, to type a function application \rle{var}, we first take
\(\dnt{\hasty{\Gamma}{1}{a}{A}{q}}\), which is guarnateed to be central, lift it
to the appropriate category \(\mc{C}_p\), and then simply postcompose it with
the denotation of \(f\). \rle{pair} is handled by composing the context split
\(\csplits{\Gamma}{\Delta}{\Xi}\) taking the tensor product of the components
\(\dnt{\hasty{\Delta}{1}{a}{A}{q}}\), \(\dnt{\hasty{\Xi}{1}{b}{B}{q}}\); this is
well-defined since both components are central and \(\mc{C}_1\) is monoidal. We
then lift the result to the appropriate \(\mc{C}_p\). The semantics of
\rle{unit} is simply dropping the entire context; \(\dnt{\cwk{\Gamma}{\cdot}}\),
while \rle{true}, and \rle{false} are each handled by first dropping the context
and then postcomposing with the appropriate constant; in all cases, this is
followed by lifting to \(\mc{C}_p\). Finally, \rle{let} and \rle{let2} are
handled by first splitting the context, then passing the right hand component to
the semantics of the bound expression \(\dnt{\hasty{\Xi}{1}{a}{A}{q}}\) or
\(\dnt{\hasty{\Xi}{1}{a}{A \otimes B}{q}}\), and then, lifting to \(\mc{C}_p\)
as necessary, after reassociating with \(\alpha\), passing everything to the
inner expression \(\dnt{\hasty{\Delta, \thyp{x}{A}{q}}{p}{e}{B}{q}}\) or
\(\dnt{\hasty{\Delta, \thyp{x}{A}{q}, \thyp{y}{B}{q}}{p}{e}{C}{q}}\). Note that
this inner expression is not necessarily central, unlike the bound expression!
